Depression and suicidality as costly signals of need

### 2021-04-02 Paper.rmd
added second depression option

Within anthropology, Raymond Firth was likely the first to put the focus on suicidality as a gamble to improve one's circumstances in the here and now. Based on observations that suicide attempts often followed social conflict or loss and varied substantially in their likelihood of death, he argued that a sizable subset of the suicide attempts among the Tikopia were not meant to end in death but instead were a means to elicit attention, status, and immediate reintegration into the community following negative events (Firth 1967[not original date]). Compared to suicide, however, anthropological explanations for depression are a much more recent occurrence (cite), despite frequent reports of depression or depression-like behavior in a wide range of societies (cite) and the potential for cross-cultural research to help understand depressive states (cite).

@@

Likely the most reliable finding when it comes to depression and suicidal behavior is that most cases are closely tied to major negative life events (Hammen, 2005 + others; citations for suicide). Compared to non-depressed individuals, those with depression tend to report at least twice as many instances of negative events in both community and clinical samples (Mazure, 1998) and more negative events than those with schizophrenia and bipolar depression across a wide range of studies (Paykel, 1994). In instances where depression is preceded by negative events, it's onset tends to quickly follow, often occurring within one month of the event (Kendler et al., 1999 [double check and find others]).

@@

Suicide attempts also tend to co-occur with social adversity (Hagen et al., 2008 + cite and describe), with evidence of this existing cross-culturally. Grief over the loss of a loved one (Firth; Devries et al. 2011), bullying (Alavi et al., 2017; Butler et al., 2019; Ford et al., 2017; Wang et al., 2020), and instances of social transgression (Firth; Malinowski; Syme and Hagen 2018; find Western citation to confirm) have all been reported to lead to suicidal behavior in both industrialized (confirm the transgression part) and traditional societies. Of the recorded instances of suicide attempts in the Human Relations Area Files (HRAF), 77.4% of attempts were found to be linked to times of social conflict, 88.7% were linked to fitness threats in general, and 84.9% were linked to situations in which attempters were powerless to resolve their situation alone (Syme et al., 2016). As with depression, assault is an particularly powerful risk factor for suicidal behavior based on multiple large-scale multi-country studies (Devries et al. 2011; Ellsberg et al. 2008) and more localized examinations. For example, both sexual and non-sexual assault were also associated with more self-harm and suicidal behaviors among both indigenous Sami and majority Norwegian adolescents, with those who attempted suicides also reporting more than twice the number of negative life events than those who did not (Reigstad and Kvernmo 2017).

## Depression and suicide attempts as both signals and bargaining tools 
One possible reason for the connection between depression and social adversity is that depression functions to elicit increased support from others in times when an individual cannot solve substantial fitness-relevant challenges on their own. The bargaining model of depression is based on the idea that for much of our evolutionary history the small groups humans lived in created many situations in which certain forms of cooperation were unlikely to be easily replaced by others while also allowing for harmful or unfair social relationships to develop (Hagen, 2003). In these instances, one's monopoly on one or more types of cooperation could be leveraged to change how one is treated if others value the depressed person's cooperation sufficiently.

>>>>

Within anthropology, Raymond Firth was likely the first to put the focus on suicidality as a gamble to improve one's circumstances in the here and now. Based on observations that suicide attempts often followed social conflict or loss and varied substantially in their likelihood of death, he argued that a sizable subset of the suicide attempts among the Tikopia were not meant to end in death but instead were a means to elicit attention, status, and immediate reintegration into the community following negative events (Firth 1967[not original date]). Compared to suicide, however, anthropological explanations for depression are a much more recent occurrence (cite examples), despite frequent reports of depression or depression-like behavior in a wide range of societies (cite) and the potential for cross-cultural research to help understand depressive states (cite).

Alternative depression idea:
- move Firth up before Syme et al. 2016 and tweak slighlty or reword Firth and put behind Syme in the same paragraph
- delete the previous depression stuff
- then follow with this
As with suicide, research on depression has been dominated by biomedical approaches of limited success. However, evolutionary and anthropological explanations for low mood have become more common. In general, these approaches see sadness and minor depression as ways to motivate adaptive behavior changes in response to negative outcomes in one's environment (cite Nesse and others), mainly differing in how they view major depression. While some approaches present most forms of major depression as dysfunctional (inflammation and some of Nesse), others have extended the logic on the benefits of low mood and reduced behavior to more severe instances of sadness. Of the possible benefits, the ability of depression to help solve fitness-relevant in times of social conflict is the most emphasized, with depression being suggested to help by reallocating resources with the body for problem solving (cite ARH), reducing the costs of interacting with higher status individuals (cite), or serving as a means to both signal one's need and bargain for increased support (cite).

@@

Likely the most reliable finding when it comes to depression and suicidal behavior is that most cases are closely tied to major negative life events (Hammen, 2005; Mazure, 1998; Devries et al. 2011; Ellsberg et al. 2008). Compared to non-depressed individuals, those with depression tend to report at least twice as many instances of negative events in both community and clinical samples (Mazure, 1998) and more negative events than those with schizophrenia and bipolar depression across a wide range of studies (Paykel, 1994). In instances where depression is preceded by negative events, it's onset tends to quickly follow, often occurring within one month of the event (Kendler et al., 1999 [double check and find others]).

@@

Suicide attempts also tend to co-occur with social adversity. Grief over the loss of a loved one (Firth; Devries et al. 2011), bullying (Alavi et al., 2017; Butler et al., 2019; Ford et al., 2017; Wang et al., 2020), and instances of social transgression (Firth; Malinowski; Syme and Hagen 2018; find Western citation to confirm) have all been reported to lead to suicidal behavior in both industrialized (confirm the transgression part) and traditional societies. Of the recorded instances of suicide attempts in the Human Relations Area Files (HRAF), 77.4% of attempts were found to be linked to times of social conflict, 88.7% were linked to fitness threats in general, and 84.9% were linked to situations in which attempters were powerless to resolve their situation alone (Syme et al., 2016). As with depression, assault is an particularly powerful risk factor for suicidal behavior based on multiple large-scale multi-country studies (Devries et al. 2011; Ellsberg et al. 2008) and more localized examinations. For example, both sexual and non-sexual assault were also associated with more self-harm and suicidal behaviors among both indigenous Sami and majority Norwegian adolescents, with those who attempted suicides also reporting more than twice the number of negative life events than those who did not (Reigstad and Kvernmo 2017).

## Depression and suicide attempts as both signals and bargaining tools

One possible reason for the connection between depression and social adversity is that depression functions to elicit increased support from others in times when an individual cannot solve substantial fitness-relevant challenges on their own. The bargaining model of depression is based on the idea that for much of our evolutionary history the small groups humans lived in created many situations in which certain forms of cooperation were unlikely to be easily replaced by others while also allowing for harmful or unfair social relationships to develop (Hagen, 2003). In these instances, one's monopoly on one or more types of cooperation could be leveraged to change how one is treated if others value the depressed person's cooperation sufficiently.
